---
title: "Class 9: Halloween Mini Project"
author: "Emily Hickey (PID: A15575724"
format: html
---

## Importing candy data

```{r}
read.csv("candy-data.csv")
```

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file,row.names=1)
head(candy)
```
>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## What is your favorite candy?

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way",]$winpercent
```
The value for my favorite candy, Milky Way,is 73.09956

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
The winpercent value for Kit Kat is 76.7686

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
The winpercent value for Tootsie Roll Snack Bars is 49.6535

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the winpercent value looks to be on a different scale to the majority of the other columns in the dataset. It was the only only that had values greater than 1.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy$chocolate
```

The zeros and ones represent whether the candy contains chocolate or not. 1 means it does, 0 means it does not.

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent) +
  geom_histogram(binwidth=10)
```

>Q9. Is the distribution of winpercent values symmetrical?

No.

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
It's above 50%. It is 50.31676

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds,12]
choc.win
```

```{r}
mean(choc.win)
```

```{r}
fruit.inds <- as.logical(candy$fruity)
fruit.win <- candy[fruit.inds,12]
fruit.win
```

```{r}
mean(fruit.win)
```
Chocolate is ranked higher than fruit candy.

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win,fruit.win)
```

Yes, they are statistical different because the p-value is 2.871e-08, which is less than .05. 

## Overall Candy Rankings 

>Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>%
  arrange(winpercent) %>% head(5)
```

Least liked: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
library(dplyr)
candy %>%
  arrange(winpercent) %>% tail(5)
```

Most liked: Reese's Peanut Butter Cup, Reese's Miniatures, Twix, Kit Kat, Snickers

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```


```{r}
my_cols=rep("black", nrow(candy))

my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols
```

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```


>Q17. What is the worst ranked chocolate candy?

The worse ranked chocolate candy is Sixlets. 

>Q18. What is the best ranked fruity candy?

The best ranked fruit candy is Starburst.

## Taking a look at pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=3.3, max.overlaps=23)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord, c(11,12)], n=5)
```

Reese's Miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord, c(11,12)], n=5)
```

Top 5 Most Expensive: Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate.
Least popular: Nik L Nip

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruit are anti-correlated.

>Q23. Similarly, what two variables are most positively correlated?

Winpercent and chocolate. 

## Principal Component Analysis 

```{r}
pca <- prcomp(candy,scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 100)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are all picked up strongly by PC1 in the positive direction. This makes sense because fruity candy does typically come in hard form and in a bag or box of multiple candies. These traits follow the correlation structure we made earlier. 