---
title: "Class 5 Data Visualization with ggplot2"
author: "Emily (PID: A15575724)"
format: pdf
---

##Using GGPLOT

The ggplot package needs to be installed as it does not come with R "out of the box."

We use the `install.packages()` function to do this.

```{r}
head(cars)
```

To use ggplot I need to load it up before i can call any of the functons in the package. I do this with the `library()` function

```{r}
library(ggplot2)
ggplot()
```

All ggplot figures have at least 3 things:
- data (the stuff we want to plot)
- aesthetic mapping (aes values)
- geoms

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() 
```

ggplot is not the only graphing system in R there are lots of others. There is even "base R" graphics


```{r}
plot(cars)
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() + 
  labs(title= "Speed and Stopping Distances of Cars", 
       x="Speed (MPH)", 
       y= "Stopping Distance (ft)",
       subtitle = "Your informative subtitle text here", 
       caption = "Dataset: 'cars'") + 
  geom_smooth(method="lm",se=FALSE) + 
  theme_bw()
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
```
```{r}
colnames(genes)
```

```{r}
ncol(genes)
```
```{r}
table(genes$State)
```

```{r}
round(table(genes$State)/nrow(genes) *100, 2)
```

```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() 
```

```{r}
p
```

```{r}
p + scale_colour_manual( values=c("blue","gray","red") )
```

```{r}
p + scale_colour_manual(values=c("blue","gray","red")) +
    labs(title="Gene Expresion Changes Upon Drug Treatment",
         x="Control (no drug) ",
         y="Drug Treatment")
```

```{r}
#install.packages("gapminder)
library(gapminder)
```


```{r}
#install.packages("dplyr")
```
```{r}
# install.packages("dplyr")  ## un-comment to install if needed
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap,
      y=lifeExp,
      color=continent,
      size=pop) + 
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap,
      y=lifeExp,
      color=pop) + 
  geom_point(alpha=0.8)
```

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap,
      y=lifeExp,
      size=pop) + 
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) 
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

