---
title: "Class 10 pt2"
author: "Emily Hickey (A15575724)"
format: html
---

## Comparative structure analysis of Adenylate Kinase 

We need some packages for todays class. These include `bio3d` and `msa`. 

The `msa` package is from BioConductor. These packages focus on genomics type work and are managed by the `BiocManager` package.

Install `install.packages("BiocManager)` and then `BiocManager::install("msa")` all entered in the R "brain" console. 

>Q10. Which of the packages above is found only on BioConductor and not CRAN?

msa

>Q11. Which of the above packages is not found on BioConductor or CRAN?:

>Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?

TRUE

```{r}
library(bio3d)

aa <- get.seq("1ake_A")
aa
```

Now I can search the PDB for related sequences: 
```{r}
b <- blast.pdb(aa)
```

```{r}
hits <- plot(b)
```

These are the related structures in the PDB that we found via BLAST search...
```{r}
head(hits$pdb.id)
```

Side-note: Lets annotate these structures (in other words find out what they are, what species they are from, stuff about the experiment they were solved in etc.)

For this we can use the `pdb.annotate()`

```{r}
anno <- pdb.annotate(hits$pdb.id)
```

```{r}
#attributes(anno)
head(anno)
```

Now we can download all these structures for further analysis with the `get.pdb()` function.
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

## Align and superpose structures
Now we have all these related structures we can Align and Superpose...

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
pdbs
```
## Principal Component Analysis

```{r}
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

